---
title: "EVST 496 Census Data"
subtitle: "Census Comparison of Holyoke, MA and Chelsea, MA and Appending"
author: "Eduardo Marin"
date: "`r format(Sys.time())`"
output:
  pdf_document:
    toc: true
  html_document:
    theme: flatly
    code_folding: hide
    fig_width: 8
    fig_height: 7
    fig_caption: true
    toc: true
    toc_float: true
    self_contained: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

# 0. Loading the packages

```{r 1. Loading Packages, message=FALSE, include=FALSE}
# 1. Loading all packages

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,   # <--- this suppresses warnings
  message = FALSE    # <--- this suppresses messages
)

packs <-c(
            'janitor'    # cleans things up, also pipe-friendly cross-tabulations
           , 'sf'         # for spatial data support
          , 'tidyverse'  # cuz
          # , 'tidylog'    # prints out what was done in dplyr and tidyr
          , 'magrittr'   # for the pipe
          , 'mapview'    # web maps for zooming and panning around
          #, 'beepr'      # makes noise when things are done!
          , 'tictoc'     # timing things.
          , 'raster'
          # , 'doParallel' # does what is says! PARALLEL
          # 'broom.mixed',# tidiers for mixed models AND nlme::gls()
          # , 'lubridate'   # DATES!
          , 'leaflet' #creating the interactive mapping elements (more specific)
          # , 'leafsync'  # linked maps
          # , 'RColorBrewer'
          )     

#2. If the packages in 'packs' are not already installed, install them
if (length(setdiff(packs, rownames(installed.packages()))) > 0) {
install.packages(setdiff(packs, rownames(installed.packages())))
}
vapply(packs, library, character.only = TRUE, logical(1), logical.return = TRUE, quietly = TRUE)
```

# 1. Read-in subset, reshape, etc.

```{r}
# 1.  Reading in the data and double checks
# Includes the subsetting of the data here (NOT IN SCRIPT 2); 
# - Have one that is a behemoth
# - Have on that is a lot smaller



# Right here is where you grab all the columns

# Right here, there will be some double checks, missingness, some quick summaries (sparkling clean dataset with only the things I need)

```

# 2. Q1 Difference in Survivorship.

## A overall

```{r}

# - We want to only look at this 25% that we have both T1 and T2
# - 1. Going to do a Tabyl on status and time (Alive vs. Dead)

```

Chi-square test here

|     | Alive              | Dead              |
|-----|--------------------|-------------------|
| T1  | number of alive t1 | number of dead t1 |
| T2  | number of alive t2 | number of dead t2 |

We can run a chi square for that above, we are testing dependence, is
there truly a difference3

```{r}
# 2. We also want a table that is per number and percent
#     Pretty version of above for reporting with a percentage basis. How do they vary in between front and backyard

```

## B Front and Backyard

```{r}

# - We want to only look at this 25% that we have both T1 and T2 and back yard
# - 1. Going to do a Tabyl on status and time (Alive vs. Dead)

```

Chi-square test here

| Time | Type  | Dead           | Alive            |
|------|-------|----------------|------------------|
| T1   | Front | number of dead | number of alive  |
| T1   | Back  | number of dead | number of alive  |
| T2   | Front | number of dead | number of alive  |
| T2   | Back  | number of dead | nnumber of alive |

We can run a chi square for that above, we are testing dependence, is
there truly a difference3

```{r}
# 2. We also want a table that is per number and percent
#     Pretty version of above for reporting with a percentage basis. How do they vary in between front and backyard

```

## C City

```{r}
# Is there a difference in survivorship across city? 

# - We want to only look at this 25% that we have both T1 and T2
# - 1. Going to do a Tabyl on status and time (Alive vs. Dead)

```

Chi-square test here

| Time | City    | Dead           | Alive            |
|------|---------|----------------|------------------|
| T1   | Chelsea | number of dead | number of alive  |
| T1   | Holyoke | number of dead | number of alive  |
| T2   | Chelsea | number of dead | number of alive  |
| T2   | Holyoke | number of dead | nnumber of alive |

We can run a chi square for that above, we are testing dependence, is
there truly a difference3

```{r}
# 2. We also want a table that is per number and percent
#     Pretty version of above for reporting with a percentage basis. How do they vary in between front and backyard

```

## D Front-back and City Interaction Effect

```{r}
# Is there a difference in survivorship across city and front and back? 

# - We want to only look at this 25% that we have both T1 and T2
# - 1. Going to do a Tabyl on status and time (Alive vs. Dead)
# -- Instaed of tabyl, potentially use group_by

```

Chi-square test here

| Time | City    | Front-back | Dead           | Alive            |
|------|---------|------------|----------------|------------------|
| T1   | Chelsea | Front      | number of dead | number of alive  |
| T1   | Chelsea | Back       | number of dead | number of alive  |
| T1   | Holyoke | Front      | number of dead | number of alive  |
| T1   | Holyoke | Back       | number of dead | nnumber of alive |
| T2   | Chelsea | Front      | number of dead | number of alive  |
| T2   | Chelsea | Back       | number of dead | number of alive  |
| T2   | Holyoke | Front      | number of dead | number of alive  |
| T2   | Holyoke | Back       | number of dead | nnumber of alive |

\#

We can run a chi square for that above, we are testing dependence, is
there truly a difference3

```{r}
# 2. We also want a table that is per number and percent
#     Pretty version of above for reporting with a percentage basis. How do they vary in between front and backyard
```

# 3. Q2 If there is a difference of survivorship between front and backyard, is the extent in which affected by median-household income?

```{r}
# The tool here is a mixed effect logistic regression, with a random effect for treeid and home value as a predictor

# mod <- lme4::lmer(status ~ yard*city + home_value + (1 | treeid))

# ^ this is a mixed effect approach

```

Another way to do what it was earlier ^ is through a survival analysis

```{r}
# It is called time to event survival analysis. This is important. We have to switch the code to dead 1 being an event while no event (aka surviving) remains 0. 

# What the survivor anlaysis is saying, like a logistic regression but accounting for time and we can put in multiple predictors. Cogs proportional hazard model
```




